---
title: 'Step 1: Harmonization of input data'
output: html_document
date: "2025-03-05"
---

```{r setup, include=FALSE, eval=TRUE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
source('0.libraries.R')
```


## Retrieve mutations' functional impact 


The mutations dataframe (somatic_mutation_dataframe.tsv) was prepared for subsequent annotation using MafAnnotator from OncoKB, as described here: https://github.com/oncokb/oncokb-annotator#oncokb-api. A
fter annotation, the processed mutation file was reimported and transformed into a 
gene-mutation matrix (Brca_mutations.tsv).
 


```{r mutations, eval=FALSE}

# Load and preprocess the somatic mutation file for OncoKB annotation
somatic_mutations <- read_tsv("somatic_mutation_dataframe.tsv") %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "Tumor_Sample_Barcode") %>%
  slice(-1) %>%  
  setNames(c("Tumor_Sample_Barcode", "Hugo_Symbol", "Variant_Classification", "HGVSp_Short")) %>%
  relocate(Hugo_Symbol, Variant_Classification, Tumor_Sample_Barcode, HGVSp_Short) %>%
  mutate(HGVSp = HGVSp_Short)  

write_delim(somatic_mutations, "somatic_mutation_3_maf.txt", delim = "\t", col_names = TRUE)



# Load the OncoKB-annotated mutation file and clinical data
somatic_mutations_annotated <- read_tsv("somatic_mutation_3_annotated_maf.txt")
clinical_dataframe <- read_tsv("clinical_dataframe.tsv")

# Process the clinical dataframe: transpose and extract relevant columns
full_clinical_data <- clinical_dataframe %>%
  column_to_rownames(var = "Name") %>%
  t() %>%
  as.data.frame() %>%
  select(5, 8) %>%  
  rownames_to_column(var = "Patient_ID") %>%
  relocate(Patient_ID) %>%
  select(-2)  

# Process somatic mutations and extract Patient ID
somatic_mutations_processed <- somatic_mutations_annotated %>%
  mutate(Patient_ID = substr(Tumor_Sample_Barcode, 1, 9) %>% str_remove("\\.")) %>%
  relocate(Patient_ID)

# Merge mutations with clinical data
mutations_clinical_merged <- left_join(somatic_mutations_processed, full_clinical_data, by = "Patient_ID")

# Filter for oncogenic mutations
oncogenic_mutations <- mutations_clinical_merged %>%
  filter(ONCOGENIC %in% c("Oncogenic", "Likely Oncogenic"))

# Create gene-mutation matrix
genes <- unique(oncogenic_mutations$Hugo_Symbol)
patients <- unique(oncogenic_mutations$Patient_ID)

# Initialize matrix with 0
genes_matrix <- matrix(0, nrow = length(patients), ncol = length(genes), 
                       dimnames = list(patients, genes))



# Map values directly onto the matrix without using a nested for-loop
oncogenic_mutations %>%
  mutate(value = ifelse(MUTATION_EFFECT %in% c("Gain-of-function", "Likely Gain-of-function"), 1, -1)) %>%
  rowwise() %>%
  mutate(genes_matrix[Patient_ID, Hugo_Symbol] <<- value)

# Convert to dataframe and save
matrice_geni <- as.data.frame(genes_matrix) %>%
  rownames_to_column(var = "Patient_ID")

write_tsv(matrice_geni, "Brca_mutations.tsv")
write_xlsx(matrice_geni, "Table_1.xlsx")


```



## PatientProfiler Step 1: Harmonization of input data 

In this first step, we processed the input breast cancer data (transcriptomics, proteomics, and phosphoproteomics) through parsing, filtering, imputation, quality control, and z-scoring.

```{r Step1, eval=FALSE}

transcriptomics_data <- read_tsv("Brca_transc.tsv") 
transcriptomics_data <- transcriptomics_data %>% dplyr::rename(gene_name = Name)

proteomics_data <- read_tsv("Brca_prot.tsv") 
proteomics_data <- proteomics_data %>% dplyr::rename(gene_name = Name) %>% select(-Database_ID)

phosphoproteomics_data <- read_tsv("Brca_phospho.tsv")  
phosphoproteomics_data <-phosphoproteomics_data %>% dplyr::rename(gene_name = Name) %>% select(-Database_ID)


omics_update(
  df_tr = transcriptomics_data,   
  df_pr = proteomics_data,   
  df_ph = phosphoproteomics_data,  
  threshold = 80,
  sw_len = 7,  
  pep_col_name = "Peptide",
  imp_method = "norm",   
  zscore = TRUE,   
  zmethod = "column",   
  metric = "median",
  output_dir = './'
)

```



## Expression level of known biomarkers

Protein expression level of known biomarkers in different subtypes, after data manipulation (Step1) (ESR1: estrogen receptor, PGR: progesterone receptor, ERBB2/HER2: human epidermal growth factor receptor 2, TOP2A: proliferative biomarker)


The result of this analysis is reported in **Figure 2F.**


```{r Expression level of known biomarkers, eval = TRUE}


clinical_dataframe <- read_tsv("clinical_dataframe.tsv") %>%
  column_to_rownames("Name") %>%
  t() %>%
  as.data.frame() %>%
  dplyr::select(Stage, NMF.Cluster) %>%
  mutate(patient_ID = rownames(.))


proteomics <- read_tsv("Proteomics_updated.tsv") %>%
  dplyr::select(-1)

data_long_proteo <- proteomics %>%
  pivot_longer(cols = -gene_name, names_to = "patient_ID", values_to = "z_score")

table_summary_proteo <- merge(clinical_dataframe, data_long_proteo, by = "patient_ID")

# Forza l'ordine dei livelli
NMF.Cluster <- c("Basal-I", "HER2-I", "LumA-I", "LumB-I")
table_summary_proteo$NMF.Cluster <- factor(table_summary_proteo$NMF.Cluster, levels = NMF.Cluster)

# 🎨 Colori e setup
my_colors <- c("Basal-I" = "brown", "HER2-I" = "darkgoldenrod", 
               "LumB-I" = "chartreuse4", "LumA-I" = "deepskyblue4")

known_biomarkers <- c("ESR1", "PGR", "ERBB2", "TOP2A")
plot_list <- list()

comparisons <- list(
  c("Basal-I", "HER2-I"),
  c("HER2-I", "LumA-I"),
  c("LumA-I", "LumB-I"),
  c("Basal-I", "LumA-I"),
  c("HER2-I", "LumB-I"),
  c("LumB-I", "Basal-I")
)

for (gene_i in known_biomarkers) {
  cat("Processing:", gene_i, "\n")
  
  table_summary_gene <- table_summary_proteo %>% filter(gene_name == gene_i)
  table_summary_gene$NMF.Cluster <- factor(table_summary_gene$NMF.Cluster, levels = NMF.Cluster)
  
  if (nrow(table_summary_gene) >= length(unique(table_summary_gene$NMF.Cluster))) {
    
    res_aov <- aov(z_score ~ NMF.Cluster, data = table_summary_gene)
    
    if (summary(res_aov)[[1]][["Pr(>F)"]][1] < 0.05) {
      
      tukey_df <- TukeyHSD(res_aov)$NMF.Cluster %>%
        as.data.frame() %>%
        rownames_to_column("comparison") %>%
        separate(comparison, into = c("group1", "group2"), sep = "(?<=-I)-") %>%
        mutate(
          group1 = as.character(group1),
          group2 = as.character(group2),
          Signif = case_when(
            `p adj` < 0.001 ~ "***",
            `p adj` < 0.01 ~ "**",
            `p adj` < 0.05 ~ "*",
            TRUE ~ "-"
          ),
          comp_key = map2_chr(group1, group2, ~ paste(sort(c(.x, .y)), collapse = "_"))
        )
      
      comparison_keys <- map_chr(comparisons, ~ paste(sort(.x), collapse = "_"))
      
      annotations <- map_chr(comparison_keys, function(key) {
        match_row <- tukey_df %>% filter(comp_key == key)
        if (nrow(match_row) == 1) as.character(match_row$Signif[[1]]) else "-"
      })
      
      plot_list[[gene_i]] <- ggplot(table_summary_gene, aes(x = NMF.Cluster, y = z_score, fill = NMF.Cluster)) +
        ggtitle(gene_i) +
        geom_boxplot() +
        scale_fill_manual(values = my_colors) +
        theme_classic() +
        geom_signif(comparisons = comparisons,
                    map_signif_level = FALSE,
                    annotations = annotations,
                    y_position = seq(3, 5, length.out = length(comparisons)))
    }
  }
}

if (length(plot_list) > 0) {
  combined_plots <- plot_grid(plotlist = plot_list, labels = "AUTO", 
                              nrow = ceiling(length(plot_list) / 2), ncol = 2)
  print(combined_plots)
}


```

