---
title: "Cross-validation"
output: html_document
date: "2025-06-11"
---

# Libraries
```{r}
knitr::opts_chunk$set(echo = T, warning = FALSE, message = FALSE)
suppressMessages(suppressWarnings(source('0.libraries.R')))
suppressMessages(suppressWarnings(source('0.test_signature_on_dataset_PP.R')))
```

# Associate patients to communities
```{r}
fold_comm <- list.files('../Step4/output_communities/')
comm_pat <- list()
for(i_fold in 1:7){
  comm_pat[[i_fold]] <- read_tsv(paste0('../Step4/Communities/output_communities/community_', 
                                        i_fold, '/patients.txt'), col_names = 'patient_id') %>% pull(patient_id)
}

names(comm_pat) <- paste0('CL', 1:7)
```

# Prepare crossvalidation input
To run the crossvalidation, for each community, sample the 70% of patients as training set (the set of patients used to extract signatures) and 30% of patients as test set. For each random sampling, we generate two transcriptomics files containing only patients of training set of each community (Train_CPTAC_coding_{i}.tsv) and only test set (Test_CPTAC_coding_{i}.tsv). 

The generated data and results for the following analyses cannot be stored on github due to size limits.
To obtain the data, feel free to contact us at veronica.lombardi@uniroma1.it and livia.perfetto@uniroma1.it.
```{r eval = F}
for (i in 1:50) {
  message("Processing iteration ", i)
  all_patients_train <- c()
  all_patients_test <- c()

  for (comm_name in names(comm_pat)) {
    patients <- comm_pat[[comm_name]]
    n <- length(patients)
    n_train <- floor(0.7 * n)

    # Sample training and test sets
    train_patients <- sample(patients, n_train)
    test_patients <- setdiff(patients, train_patients)
    all_patients_train <- c(all_patients_train, train_patients)
    all_patients_test <- c(all_patients_test, train_patients)

    # Create output folders
    train_dir <- file.path(paste0("Crossvalidation_input/output_communities_train_", i), comm_name)
    test_dir  <- file.path(paste0("Crossvalidation_input/output_communities_test_", i), comm_name)

    dir.create(train_dir, recursive = TRUE, showWarnings = FALSE)
    dir.create(test_dir, recursive = TRUE, showWarnings = FALSE)

    # Write to file
    writeLines(train_patients, file.path(train_dir, "patients.txt"))
    writeLines(test_patients, file.path(test_dir, "patients.txt"))
  }

  tr_df <- read_tsv("../Step5/Transcriptomics_updated_CPTAC_coding.tsv")
  tr_df_train <- tr_df %>% select(gene_name, all_patients_train)
  tr_df_test <- tr_df %>% select(gene_name, all_patients_test)
  write_tsv(tr_df_train, paste0('Transcriptomics_data_train_test/Train_CPTAC_coding_', i, '.tsv'))
  write_tsv(tr_df_test, paste0('Transcriptomics_data_train_test/Test_CPTAC_coding_', i, '.tsv'))
}
```

# Run crossvalidation 

## I) Extract signatures from patients subsets

```{r eval = F}
library(future.apply)
plan(multisession, workers = 10)

future_lapply(1:50, function(i) {
  
  source('./0.libraries.R')
  
  message("Running iteration ", i)
  
  #Run for communities 2, 3,4 , 7
  extract_signatures(base_path = paste0("./Crossvalidation_input/output_communities_train_", i, "/"),
                     transcriptomics_file =  paste0('Transcriptomics_data_train_test/Train_CPTAC_coding_', i, '.tsv'),
                     output_dir = paste0("./Crossvalidation_signatures/Run_for_communities_2_3_4_7_training_", i),
                     diff_thres = 0.7)
  
  #Run for communities 1, 5, 6
  extract_signatures(base_path =  paste0("./Crossvalidation_input/output_communities_train_", i, "/"),
                     transcriptomics_file = paste0('Transcriptomics_data_train_test/Train_CPTAC_coding_', i, '.tsv'),
                     output_dir = paste0("./Crossvalidation_signatures//Run_for_communities_1_5_6_training_", i),
                     diff_thres = 0.5)
  
  return(NULL)
})
```

Combine the signatures in a single folder. 
```{r eval = F}
for(i in c(1:50)){
  # Define source directories
  parent_dir_1 <- paste0("./Crossvalidation_signatures/Run_for_communities_1_5_6_training_", i)
  parent_dir_2 <- paste0("./Crossvalidation_signatures/Run_for_communities_2_3_4_7_training_", i)

  dir_1 <- file.path(parent_dir_1, "Signatures")
  dir_2 <- file.path(parent_dir_2, "Signatures")

  # Define the files to extract
  files_1 <- c("signature_community_1.tsv", "signature_community_5.tsv", "signature_community_6.tsv")
  files_2 <- c("signature_community_2.tsv", "signature_community_3.tsv", "signature_community_4.tsv", "signature_community_7.tsv")

  # Define the destination directory
  dest_dir <- paste0("./Crossvalidation_signatures/Run_for_communities_training_", i)
  dir.create(dest_dir)

  # Function to move files
  move_files <- function(file_list, source_dir, destination_dir) {
    for (file in file_list) {
      source_path <- file.path(source_dir, file)
      dest_path <- file.path(destination_dir, file)

      # Check if the file exists before moving
      if (file_exists(source_path)) {
        file_move(source_path, dest_path)
        cat("Moved:", file, "to", destination_dir, "\n")
      } else {
        cat("File not found:", file, "\n")
      }
    }
  }

  # Move the required files
  move_files(files_1, dir_1, dest_dir)
  move_files(files_2, dir_2, dest_dir)

  # Delete the directories, including their parent folders
  if (dir_exists(parent_dir_1)) {
    dir_delete(parent_dir_1)
    cat("Deleted directory:", parent_dir_1, "\n")
  }
  if (dir_exists(parent_dir_2)) {
    dir_delete(parent_dir_2)
    cat("Deleted directory:", parent_dir_2, "\n")
  }
}
```

## II) Run FGSEA analysis on CPTAC validation data
```{r eval = F}
source('0.run_fgsea_signatures.R')

library(future)
library(future.apply)
library(readr)

# Set up parallel plan (adjust depending on your system: multisession for Windows, multicore for Unix-like)
#plan(multisession)  # or plan(multicore) on Linux/macOS
plan(multisession, workers = 10)

# Define the function to apply in parallel
run_analysis <- function(i_run) {
  dest_dir <- paste0("./Crossvalidation_signatures/Run_for_communities_training_", i_run)
  
  n_com <- list.files(dest_dir)
  
  gene_signatures <- list()
  for (i in seq_along(n_com)) {
    comm_f <- read_tsv(paste0(dest_dir, '/signature_community_', i, '.tsv'), show_col_types = FALSE)
    gene_signatures[[i]] <- comm_f$gene_name
  }
  names(gene_signatures) <- paste0("Comm_", seq_along(n_com))
  
  transcriptomics_data <- read_tsv(paste0('Transcriptomics_data_train_test/Test_CPTAC_coding_', i_run, '.tsv'))
  
  enrichment_result <- tryCatch(
    {
      run_fgsea_signatures(
        gene_signatures = gene_signatures,
        transcriptomic_data = transcriptomics_data,
        pval_cutoff = 0.05,
        NES_filter = TRUE
      )
    },
    error = function(e) {
      message("Error in run ", i_run, ": ", e$message)
      return(NULL)
    }
  )
  
  if (!is.null(KM_list)) {
    write_tsv(enrichment_result, paste0('./BRCA_enrichment_test/fgsea_cptac_test_', i_run, '.tsv'))
  }
  
  return(NULL)  # return nothing or custom status
}

# Run in parallel over 1:50
future_lapply(1:50, run_analysis)
```

# Compute key metrics

## Precision, Recall and F1 score
```{r eval = F}

# Get true labels once (assuming static across runs)
com_TP <- stack(comm_pat)
colnames(com_TP) <- c("patient_id", "community")

# Initialize list to collect metrics
all_metrics <- list()

for (i_run in 1:50) {
  
  # Read enrichment and transcriptomics data
  enrichment_df <- read_tsv(paste0("./BRCA_enrichment_test/fgsea_cptac_test_", i_run, ".tsv"), show_col_types = FALSE)
  transcriptomics_data <- read_tsv(paste0('Transcriptomics_data_train_test/Test_CPTAC_coding_', i_run, '.tsv'), 
                                   show_col_types = FALSE)
  test_patients <- colnames(transcriptomics_data)[-1]

  # Infer community label from enrichment
  sign_pats <- enrichment_df %>%
    filter(NES > 0) %>%
    group_by(patient_id) %>%
    slice_min(order_by = padj, n = 1) %>%
    slice_max(order_by = NES, n = 1) %>%
    ungroup() %>%
    filter(padj < 0.01) %>%
    dplyr::rename('inferred_label' = 'pathway') %>%
    mutate(inferred_label = str_replace(inferred_label, "^Comm_(\\d+)$", "CL\\1"))
  
  # Join true and inferred labels
  com_TP_inf <- full_join(sign_pats, com_TP, by = 'patient_id') %>%
    filter(patient_id %in% test_patients)

  # Compute metrics
  metrics <- com_TP_inf %>%
    dplyr::group_by(community) %>%
    dplyr::summarise(
      TP = sum(community == inferred_label, na.rm = TRUE),
      FN = sum(is.na(inferred_label) & !is.na(community)),
      FP = sum(community != inferred_label, na.rm = TRUE),
      Precision = ifelse((TP + FP) > 0, TP / (TP + FP), NA),
      Recall = ifelse((TP + FN) > 0, TP / (TP + FN), NA),
      F1 = ifelse(!is.na(Precision) & !is.na(Recall) & (Precision + Recall) > 0,
                   2 * Precision * Recall / (Precision + Recall), NA),
      .groups = "drop"
    ) %>%
    mutate(run = i_run)
  
  # Store results
  all_metrics[[i_run]] <- metrics
}

# Combine all metrics into one dataframe
final_metrics<- bind_rows(all_metrics)

write_tsv(final_metrics, './results/crossvalidation_f1_score.tsv')
```

## Patient community size and signatures' size

### Patients' number 
```{r eval  = F}
for (i in 1:50) {
  # Create output folders
  train_dir <- file.path(paste0("Crossvalidation_input/output_communities_train_", i))
  test_dir  <- file.path(paste0("Crossvalidation_input/output_communities_test_", i))
  
  #i_comm = 1
  for(i_comm in c(1:7)){
    comm_pat <- read_tsv(paste0(train_dir, '/community_', i_comm, '/patients.txt'), col_names = c('patient_id'))
    comm_pat$community <- paste0('CL', i_comm)
    comm_pat$run <- paste0('run_', i)
    comm_pat$group <- 'train'
    
    comm_pat_t <- read_tsv(paste0(test_dir, '/community_', i_comm, '/patients.txt'), col_names = c('patient_id'))
    comm_pat_t$community <- paste0('CL', i_comm)
    comm_pat_t$run <- paste0('run_', i)
    comm_pat_t$group <- 'test'
    
    comm_pat_i <- bind_rows(comm_pat, comm_pat_t)
    if(i == 1 & i_comm == 1){
      comm_pat_all <- comm_pat_i
    }else{
      comm_pat_all <- bind_rows(comm_pat_all, comm_pat_i)
    }
  }
}

write_tsv(comm_pat_all, './results/crossvalidation_patients_dim.tsv')
```

### Signatures' size
```{r eval = F}
for (i_run in 1:50) {
  #i_run = 1
  dest_dir <- paste0("./Crossvalidation_signatures/Run_for_communities_training_", i_run)
  
  n_com <- list.files(dest_dir)
  gene_signatures <- list()
  for (i in seq_along(n_com)) {
    comm_f <- read_tsv(paste0(dest_dir, '/signature_community_', i, '.tsv'), show_col_types = FALSE)
    gene_signatures[[i]] <- comm_f$gene_name
  }
  names(gene_signatures) <- paste0("CL", seq_along(n_com))
  
  gene_signatures_df <- stack(gene_signatures)
  gene_signatures_df$run <- paste0('run_', i_run)
  if(i_run == 1){
    gene_signatures_df_all <- gene_signatures_df
  }else{
    gene_signatures_df_all <- bind_rows(gene_signatures_df_all, gene_signatures_df)
  }
}

gene_signatures_df_all %>% 
  dplyr::count(ind, run) %>%
  dplyr::group_by(ind) %>%
  dplyr::summarise(dim_sign = round(mean(n),0)) -> signatures_average_dim

write_tsv(signatures_average_dim, './results/crossvalidation_signatures_dim.tsv')
```

### Intersection and Jaccard index
```{r eval = F}
jaccard_results <- list()

for (i_run in 1:50) {
  dest_dir <- paste0("./Crossvalidation_signatures/Run_for_communities_training_", i_run)
  
  n_com <- list.files(dest_dir)
  
  gene_signatures <- list()
  for (i in seq_along(n_com)) {
    comm_f <- read_tsv(paste0(dest_dir, '/signature_community_', i, '.tsv'), show_col_types = FALSE)
    gene_signatures[[i]] <- comm_f$gene_name
  }
  names(gene_signatures) <- paste0("CL", seq_along(n_com))
  
  signatures_genes <- stack(gene_signatures)
  colnames(signatures_genes) <- c("gene_name", "community")
  signatures_genes$run <- paste0('run_', i_run)
  
  if(i_run == 1){
    all_signatures <- signatures_genes
  }else{
    all_signatures <- bind_rows(all_signatures, signatures_genes)
  }
}

# Get patientprofiler signatures
dest_dir <- '../Step5'
gene_signatures <- list()

for (i in seq_along(n_com)) {
    comm_f <- read_tsv(paste0(dest_dir, '/signature_community_', i, '.tsv'), show_col_types = FALSE)
    gene_signatures[[i]] <- comm_f$gene_name
    }

names(gene_signatures) <- paste0("CL", seq_along(n_com))
  
patientprofiler_genes <- stack(gene_signatures)
colnames(patientprofiler_genes) <- c("gene_name", "community")
patientprofiler_genes$run <- 'patientprofiler'

pairs <- unique(all_signatures[, c("run", "community")])

for (i in seq_len(nrow(pairs))) {
  run_i <- pairs$run[i]
  community_i <- pairs$community[i]
  
  # Get gene sets
  genes_set <- all_signatures %>%
    filter(run == run_i, community == community_i) %>%
    pull(gene_name) %>%
    unique()
  
  signature_set <- patientprofiler_genes %>%
    filter(community == community_i) %>%
    pull(gene_name) %>%
    unique()
  
  # Compute Jaccard Index safely
  union_len <- length(union(genes_set, signature_set))
  intersection_len <- length(intersect(genes_set, signature_set))
  
  jaccard <- if (union_len > 0) intersection_len / union_len else NA_real_
  
  # Store result
  jaccard_results[[i]] <- data.frame(
    run = run_i,
    community = community_i,
    jaccard_index = jaccard,
    intersection = intersection_len,
    sign_length = length(signature_set)
  )
}

# Combine all results into a single dataframe
jaccard_df <- do.call(rbind, jaccard_results)
write_tsv(jaccard_df, './results/crossvalidation_jaccard.tsv')
```


# Plots

Define colors
```{r}
group_colors <- c('CL1' = "grey",
                  'CL2' = "#59BDB2", 
                  'CL3' = "#E7DBAE", 
                  'CL4' = "#F4A2A4", 
                  'CL5' = "#E94587", 
                  'CL6' = "#9CCEE3",
                  'CL7' = "#6B2B68")

```

F1 score across runs
```{r}
final_metrics <- read_tsv('./results/crossvalidation_f1_score.tsv')
final_metrics %>% 
  ggplot(aes(x = community, y = F1, fill = community)) +
  scale_fill_manual(values = group_colors) +
  geom_boxplot() +
  theme_classic() +
  xlab('') +
  ylab('F1 score across 50 runs') +
  theme(axis.text.x = element_text(angle = 90, vjust = .5)) -> boxplot_F1score
```


Patients number
```{r}
comm_pat_all <- read_tsv('./results/crossvalidation_patients_dim.tsv')

comm_pat_all %>% 
  distinct(community, patient_id) %>% 
  dplyr::count(community) %>%
  ggplot(aes(x = community, y = n, fill = community)) +
  geom_bar(stat = 'identity') + 
  scale_fill_manual(values = group_colors) +
  xlab('') +
  ylab('# CPTAC Patients in community') +
   geom_text(aes(label = n,vjust = 0)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = .5)) -> barplot_patients_comm
```

Signature avg size in crossvalidation
```{r}
signatures_average_dim <- read_tsv('./results/crossvalidation_signatures_dim.tsv')

colnames(signatures_average_dim)[1] <- 'community'

signatures_average_dim %>% 
  ggplot(aes(x = community, y = dim_sign, fill = community)) +
  geom_bar(stat = 'identity') + 
  scale_fill_manual(values = group_colors) +
  xlab('') +
  ylab('Avg signature size in \n50 runs of crossvalidation') +
  theme_classic() +
  geom_text(aes(label = dim_sign,vjust = 0)) +
  theme(axis.text.x = element_text(angle = 90, vjust = .5)) -> barplot_sign_size
```

Intersection among original and crossvalidation signatures
```{r}
jaccard_df <- read_tsv('./results/crossvalidation_jaccard.tsv')
jaccard_df %>%
  ggplot(aes(x = community, y = intersection/sign_length * 100, fill = community))+
  geom_boxplot() +
  scale_fill_manual(values = group_colors) +
  xlab('') +
  ylab('% of signatures genes found') +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = .5)) -> boxplot_intesection
```

## Supplementary Figure
```{r}

plot_list_cross <- list(boxplot_intesection, boxplot_F1score, barplot_patients_comm, barplot_sign_size)

cowplot::plot_grid(plotlist = plot_list_cross, ncol = 2, 
                   align = 'hv',
                   labels = LETTERS[1:4])
# ggsave('../Figures/crossvalidation_cowplot.pdf', crossvalidation,
#        width = 8, height = 6, dpi = 300, device = 'pdf')
```
**Supplementary Figure. Cross-validation results for community-specific classification performance.** **A)** Percentage of signature genes recovered in each of 50 cross-validation runs per community. **B)** F1 score distributions showing classification performance across runs. **C)** Number of CPTAC patients assigned to each community in the full run. **D)** Average number of genes included in the signature for each community across the 50 cross-validation runs.  





