---
title: "Robustness analysis: randomization"
output: html_document
date: "2025-06-11"
---

# Libraries
```{r}
knitr::opts_chunk$set(echo = T, warning = FALSE, message = FALSE)
suppressMessages(suppressWarnings(source('0.libraries.R')))
suppressMessages(suppressWarnings(source('0.test_signature_on_dataset_PP.R')))
```

The generated data and results for the following analyses cannot be stored on github due to size limits.
To obtain the data, feel free to contact us at veronica.lombardi@uniroma1.it and livia.perfetto@uniroma1.it.

# Shuffle CPTAC transcriptomics and enrich TCGA 

### Shuffling CPTAC transcriptomics data
```{r eval = F}
tr_df_cptac <- read_tsv("../Step5/Transcriptomics_updated_CPTAC_coding.tsv")
tr_df_cptac %>% column_to_rownames('gene_name') -> expr_matrix

set.seed(42)  # for reproducibility

# Suppose expr_matrix is your genes x samples matrix
shuffle_expr_list <- replicate(
  n = 50,
  expr_matrix %>%
    apply(1, sample) %>%  # shuffle each gene's expression values
    t(),                  # transpose back to genes x samples
  simplify = FALSE
)

patients <- colnames(tr_df_cptac)

dir.create('Shuffled_CPTAC_transcriptomics')
for(i in c(1:50)){
  tr_df_i <- shuffle_expr_list[[i]]
  genes <- rownames(tr_df_i)

  colnames(tr_df_i) <- patients[-1]
  as_tibble(tr_df_i) -> tr_df_i
  tr_df_i$gene_name <- genes
  tr_df_i <- tr_df_i %>% relocate('gene_name')

  write_tsv(tr_df_i, paste0('Shuffled_CPTAC_transcriptomics/CPTAC_transcriptomics_shuffled_', i, '.tsv'))
}
```

### Perform ANOVA on TCGA patients on multithreading 
```{r eval = F}
 library(future.apply)
plan(multisession, workers = 10)

future_lapply(1:50, function(i) {

  source('0.library.R')

  message("Running iteration ", i)

  #Run for communities 2, 3,4 , 7
  extract_signatures(base_path = "../Step4/output_communities/",
                     transcriptomics_file = paste0('Shuffled_CPTAC_transcriptomics/CPTAC_transcriptomics_shuffled_', i, '.tsv'),
                     output_dir = paste0("./Run_for_communities_2_3_4_7_shuffled_", i),
                     diff_thres = 0.7)

  #Run for communities 1, 5, 6
  extract_signatures(base_path = "../Step4/output_communities/",
                     transcriptomics_file = paste0('Shuffled_CPTAC_transcriptomics/CPTAC_transcriptomics_shuffled_', i, '.tsv'),
                     output_dir = paste0("./Run_for_communities_1_5_6_shuffled_", i),
                     diff_thres = 0.5)

  return(NULL)
})
```

Order the output of different thresholds in a new folder
```{r eval = F}
for(i in c(1:50)){
  # Define source directories
  parent_dir_1 <- paste0("./Run_for_communities_1_5_6_shuffled_", i)
  parent_dir_2 <- paste0("./Run_for_communities_2_3_4_7_shuffled_", i)

  dir_1 <- file.path(parent_dir_1, "Signatures")
  dir_2 <- file.path(parent_dir_2, "Signatures")

  # Define the files to extract
  files_1 <- c("signature_community_1.tsv", "signature_community_5.tsv", "signature_community_6.tsv")
  files_2 <- c("signature_community_2.tsv", "signature_community_3.tsv", "signature_community_4.tsv", "signature_community_7.tsv")

  # Define the destination directory
  dest_dir <- paste0("./Run_for_communities_shuffled_", i)
  dir.create(dest_dir)

  # Function to move files
  move_files <- function(file_list, source_dir, destination_dir) {
    for (file in file_list) {
      source_path <- file.path(source_dir, file)
      dest_path <- file.path(destination_dir, file)

      # Check if the file exists before moving
      if (file_exists(source_path)) {
        file_move(source_path, dest_path)
        cat("Moved:", file, "to", destination_dir, "\n")
      } else {
        cat("File not found:", file, "\n")
      }
    }
  }

  # Move the required files
  move_files(files_1, dir_1, dest_dir)
  move_files(files_2, dir_2, dest_dir)

  # Delete the directories, including their parent folders
  if (dir_exists(parent_dir_1)) {
    dir_delete(parent_dir_1)
    cat("Deleted directory:", parent_dir_1, "\n")
  }
  if (dir_exists(parent_dir_2)) {
    dir_delete(parent_dir_2)
    cat("Deleted directory:", parent_dir_2, "\n")
  }
}

```

# Run FGSEA in parallel on TCGA data
```{r eval = F}

clinic <- read_tsv("../Step5/clinical_TCGA.tsv") 

# Remove first two rows (assuming they contain metadata)
clinic <- clinic[-(1:2), ]

# Create a dataframe with patient survival information 
## V2 improvement Livia
df_clinical_status <- data.frame(
  patient_id = clinic$patient_ID,
  serial_time = as.numeric(ifelse(clinic$death_days_to == '[Not Applicable]', 
                                  clinic$last_contact_days_to, clinic$death_days_to)),
  status = ifelse(clinic$vital_status == 'Alive', 0, 1)  # Initialize status as 0 (alive)
)

# Order by survival time
clinical_data_formatted <- df_clinical_status[order(df_clinical_status$serial_time), ]

tcga_transcriptomics <- read_tsv('./transcriptomics_TCGA_zscored.tsv')


library(future)
library(future.apply)
library(readr)

# Set up parallel plan (adjust depending on your system: multisession for Windows, multicore for Unix-like)
#plan(multisession)  # or plan(multicore) on Linux/macOS
plan(multisession, workers = 10)

# Define the function to apply in parallel
run_analysis <- function(i_run) {
  dest_dir <- paste0("./Shuffled_signatures/Run_for_communities_shuffled_", i_run)
  
  n_com <- list.files(dest_dir)
  
  gene_signatures <- list()
  for (i in seq_along(n_com)) {
    comm_f <- read_tsv(paste0(dest_dir, '/signature_community_', i, '.tsv'), show_col_types = FALSE)
    gene_signatures[[i]] <- comm_f$gene_name
  }
  names(gene_signatures) <- paste0("Comm_", seq_along(n_com))
  
  KM_list <- tryCatch(
    {
      test_signature_on_dataset(
        gene_signatures = gene_signatures,
        transcriptomic_data = tcga_transcriptomics,
        clinical_data = clinical_data_formatted,
        pval_cutoff = 0.05,
        NES_filter = TRUE
      )
    },
    error = function(e) {
      message("Error in run ", i_run, ": ", e$message)
      return(NULL)
    }
  )
  
  if (!is.null(KM_list)) {
    write_tsv(KM_list$enrichment, paste0('./BRCA_enrichment/fgsea_tcga_shuffle_', i_run, '.tsv'))
  }
  
  return(NULL)  # return nothing or custom status
}

# Run in parallel over 1:50
future_lapply(1:50, run_analysis)
```

# Create a TCGA results table 
```{r eval = F}
# Read TCGA enrichment in original PatientProfiler run
tcga_subtypes <- readxl::read_excel('../Step5/Table_3.xlsx',sheet = 2)%>%
  dplyr::select(patient_id = patient_ID, subtype = Subtype, patientprofiler = Signature) %>%
  dplyr::filter(!is.na(subtype)) %>% dplyr::filter(subtype != 'Normal')

tcga_subtypes$patientprofiler <- as.numeric(gsub('Signature', '', tcga_subtypes$patientprofiler))


clinic <- read_tsv("../Step5/clinical_TCGA.tsv") 

# Remove first two rows (assuming they contain metadata)
clinic <- clinic[-(1:2), ]

# Create a dataframe with patient survival information 
## V2 improvement Livia
df_clinical_status <- data.frame(
  patient_id = clinic$patient_ID,
  serial_time = as.numeric(ifelse(clinic$death_days_to == '[Not Applicable]', 
                                  clinic$last_contact_days_to, clinic$death_days_to)),
  status = ifelse(clinic$vital_status == 'Alive', 0, 1)  # Initialize status as 0 (alive)
)

# Order by survival time
clinical_data_formatted <- df_clinical_status[order(df_clinical_status$serial_time), ]

gene_signatures <- list()
transcriptomic_data <- tibble()

pvalues_table <- tibble(run_id = character(), 
                        pvalue_KM = numeric(),
                        n_sign_KM = numeric(),
                        enriching_signatures = character())

proportions <- c(5, 10, 15, 25, 50, 75, 100)

for(prop in proportions){
  
  for (i_run in 1:50) {
     run_id <- paste0('shuffling_', prop, 'perc_', i_run)
    
    if(prop != 100){
      key = paste0(prop, 'perc_', i_run)
    }else{
      key = i_run
    }
    
    if(file.exists(paste0('./BRCA_enrichment/fgsea_tcga_shuffle_', key, '.tsv'))){
      enrichment_raw <- read_tsv(paste0('./BRCA_enrichment/fgsea_tcga_shuffle_', key, '.tsv')) 
    }else{
      next
    }
    
    enrichment <- enrichment_raw %>%
      dplyr::select(patient_id, signature = pathway)
    
    enrichment %>%
      dplyr::count(signature) %>%
      filter(n < 5) %>%
      pull(signature) -> to_exclude
    
    enrichment %>% 
      distinct(signature) %>% 
      filter(!signature %in% to_exclude) %>%
      pull(signature) -> to_keep
    
    
    n_sign_KM <- length(unique(enrichment$signature)) - length(to_exclude)
    
    # Remove communities without enough patients
    enrichment <- enrichment %>% filter(signature %in% to_keep)
    enrichment$signature <- gsub('Comm_', '', enrichment$signature)
    colnames(enrichment)[2] <- run_id
    
    tcga_subtypes <- full_join(tcga_subtypes, 
                               enrichment, by = 'patient_id')
    
    if(length(to_keep) <= 1){
      pvalues_table <- pvalues_table %>% 
        add_row(run_id = run_id, 
                pvalue_KM = NA,
                n_sign_KM = n_sign_KM, 
                enriching_signatures = paste0(to_keep, collapse = ';'))
      
      next
    }
    
    # Generate pvalue matrix 
    test_signature_on_dataset(
      gene_signatures = gene_signatures,
      transcriptomic_data = transcriptomic_data,
      clinical_data = clinical_data_formatted,
      pval_cutoff = 0.01,
      NES_filter = TRUE,
      enrichment_result = enrichment_raw,
      time_threshold = 365*5,
      exclude_communities = to_exclude
    ) -> KM_list
    
      # Create pvalue table
      pvalues_table <- pvalues_table %>% 
        add_row(run_id = run_id, 
                pvalue_KM = KM_list$adjusted_p_value,
                n_sign_KM = n_sign_KM, 
                enriching_signatures = paste0(to_keep, collapse = ';'))
  }
}

pvalues_table_adj <- pvalues_table %>%
  mutate(p_value_adj_KM = p.adjust(pvalue_KM, method = 'BH'))

write_tsv(pvalues_table_adj, './results/pvalue_table_shuffling.tsv')
write_tsv(tcga_subtypes, './results/tcga_results_shuffling.tsv')
```


# Compare control (shuffled) and original signatures
```{r eval = F}
proportions <- c(5, 10, 15, 25, 50, 75, 100)

# Read TCGA enrichment result using original signatures
cptac_results <- readxl::read_excel('../Step5/Table_3.xlsx',sheet = 1)%>%
  dplyr::select(patient_id = patient_ID, subtype = Subtype, patientprofiler = Signature) %>%
  dplyr::mutate(subtype = str_to_upper(str_remove(pattern = '\\-I', subtype)),
                patientprofiler = str_remove(pattern = 'Signature', patientprofiler)) 


# Read TCGA enrichment result using shuffled signatures
tcga_results <- read_tsv('./results/tcga_results_shuffling.tsv') %>%
  filter(!is.na(subtype)) %>%
  dplyr::mutate(subtype = str_to_upper(subtype))

max_values_cptac <- sapply(cptac_results[,-c(1:2)], max, na.rm = TRUE)

all_results <- tibble()

runs <- unlist(lapply(paste0('shuffling_', proportions, 'perc'), function(x) paste0(x, '_', 1:50)))

runs <- c('patientprofiler', runs)

cptac <- cptac_results %>% dplyr::select(subtype)


for(run in runs){
  groups <- 1:7
  
  for (group_id in groups) {
    if(run %in% colnames(tcga_results)){
      cptac <- cptac_results %>% 
        filter(.data[['patientprofiler']] == group_id) %>% 
        dplyr::select(subtype)
      tcga <- tcga_results %>% 
        filter(.data[[run]] == as.character(group_id)) %>% 
        dplyr::select(subtype)
    }else{
      cosine_similarity <- NA
      next
    }
    
    # Initialize output values
      cosine_similarity <- NA
      
      if (nrow(tcga) > 0 & nrow(cptac) > 0) {
        subtype_table <- table(
          c(tcga$subtype, cptac$subtype),
          c(rep("tcga", nrow(tcga)), rep("cptac", nrow(cptac)))
        )
        
        if (nrow(subtype_table) > 1) {
          vec_cptac <- subtype_table[,1]
          vec_tcga <- subtype_table[,2]
          cosine_similarity <- sum(vec_cptac * vec_tcga) / (sqrt(sum(vec_cptac^2)) * sqrt(sum(vec_tcga^2)))
        }
      }
      # Store result
      all_results <- bind_rows(all_results, tibble(
        run = run,
        group = group_id,
        cosine_similarity = cosine_similarity
      ))
  }
}
```


Create a table with similarity among subtypes and prognostic value
```{r eval = F}

pvalues_table <- read_tsv('./results/pvalue_table_shuffling.tsv') 

all_results %>%
  dplyr::group_by(run) %>%
  dplyr::summarise(avg_cosine = mean(cosine_similarity, na.rm = TRUE)) -> summary_results

summary_results_pval <- inner_join(summary_results, pvalues_table, 
                                   by = c('run' = 'run_id'))
summary_results_pval$signatures_n <- NULL

summary_results_pval$run <- str_remove(summary_results_pval$run, 'shuffling_')
write_tsv(summary_results_pval, './results/shuffling_result.tsv')
```

# Plots

Read results
```{r}
shuffl_result <- read_tsv('./results/shuffling_result.tsv') 
shuffl_result %>% separate(run, into = c('perc', 'iter')) -> shuffl_result

shuffl_result$perc <- gsub('perc', '%', shuffl_result$perc)

shuffl_result$perc <- factor(shuffl_result$perc, levels = c('0%', '5%', '10%', '15%', '25%', '50%', '75%', '100%'))
```

## Signatures number
```{r}
shuffl_result[-1,] %>%
  ggplot(aes(x = perc, y = n_sign_KM))+
  geom_boxplot() +
  theme_classic()+
  ylab('# of signatures') +
  xlab('% of shuffled genes') -> signatures_num
```

## Prognostic value
```{r}
# Define bins and colors (light grays)
pval_bins <- data.frame(
  ymin = c(0, 0.01, 0.05, 0.1, 0.25, 0.5),
  ymax = c(0.01, 0.05, 0.1, 0.25, 0.5, 1),
  fill = factor(c("<0.01", "0.01–0.05", "0.05–0.1", "0.1–0.25", "0.25–0.5", ">0.5"),
                levels = c("<0.01", "0.01–0.05", "0.05–0.1", "0.1–0.25", "0.25–0.5", ">0.5"))
)

# Plot with background bins
ggplot() +
  geom_rect(data = pval_bins,
            aes(ymin = ymin, ymax = ymax, xmin = -Inf, xmax = Inf, fill = fill),
            alpha = 0.2) +
  geom_boxplot(data = shuffl_result %>% filter(perc != '0%') %>% drop_na(pvalue_KM),
               aes(x = perc, y = pvalue_KM), outlier.shape = NA) +
  geom_jitter(data = shuffl_result %>% filter(perc != '0%') %>% drop_na(pvalue_KM),
              aes(x = perc, y = pvalue_KM), width = 0.2, size = .3) +
  geom_hline(yintercept = 0.05, color = 'red') +
  scale_fill_brewer(palette = "Greys", name = "P-value range", direction = -1) +
  labs(
    x = "% of shuffled genes",
    y = "Signatures groups adj p-values in Kaplan-Meier curves"
  ) +
  theme_classic() +
  theme(legend.position = "right") -> boxplot_pvalues_bins
```

## Subtypes similarity
```{r}
pp_threshold <- 0.867

shuffl_result %>% 
  ggplot(aes(x = perc, y = avg_cosine)) +
  geom_boxplot(outlier.shape = NA)+
  geom_jitter( width = 0.2, size = .3) +
   geom_hline(yintercept = 0.867, color = 'red', linetype = 'dashed') +
   labs(
    x = "% of shuffled genes",
    y = "Avg cosine similarity across signatures"
  ) +
  theme_classic() +
  ylim(0, 1) -> plot_classification
```

## Check signatures similarity
To obtain the Shuffled_signatures folder contact us.
```{r eval = F}
jaccard_results <- list()

for(prop in c(5, 10, 15, 25, 50, 75, 100)){
  for (i_run in 1:50) {
    if(prop == 100){
       dest_dir <- paste0("./Shuffled_signatures/Run_for_communities_shuffled_", i_run)
    }else{
       dest_dir <- paste0("./Shuffled_signatures/Run_for_communities_shuffled_", prop, 'perc_', i_run)
    }
   
    n_com <- list.files(dest_dir)
    gene_signatures <- list()
    for (i in seq_along(n_com)) {
      comm_f <- read_tsv(paste0(dest_dir, '/signature_community_', i, '.tsv'), 
                         show_col_types = FALSE)
      gene_signatures[[i]] <- comm_f$gene_name
      }
    names(gene_signatures) <- paste0("community_", seq_along(n_com))
  
    signatures_genes <- stack(gene_signatures)
    colnames(signatures_genes) <- c("gene_name", "community")
    signatures_genes$run <- paste0(prop, 'perc_', i_run)
    
    if(i_run == 1 & prop == 5){
      all_signatures <- signatures_genes
      }else{
      all_signatures <- bind_rows(all_signatures, signatures_genes)
    }
  }
}

# Get PatientProfiler original signatures
dest_dir <- '../Step5'
gene_signatures_pp <- list()

i = 1
for (i in seq_along(n_com)) {
    comm_f <- read_tsv(paste0(dest_dir, '/signature_community_', i, '.tsv'), show_col_types = FALSE)
    gene_signatures_pp[[i]] <- comm_f$gene_name
    }
names(gene_signatures_pp) <- paste0("community_", seq_along(n_com))
  
  patientprofiler_genes <- stack(gene_signatures_pp)
colnames(patientprofiler_genes) <- c("gene_name", "community")
  patientprofiler_genes$run <- 'patientprofiler'


pairs <- unique(all_signatures[, c("run", "community")])

i = 1
for (i in seq_len(nrow(pairs))) {
  run_i <- pairs$run[i]
  community_i <- pairs$community[i]
  
  # Get gene sets
  genes_set <- all_signatures %>%
    filter(run == run_i, community == community_i) %>%
    pull(gene_name) %>%
    unique()
  
  signature_set <- patientprofiler_genes %>%
    filter(community == community_i) %>%
    pull(gene_name) %>%
    unique()
  
  # Compute Jaccard Index safely
  union_len <- length(union(genes_set, signature_set))
  intersection_len <- length(intersect(genes_set, signature_set))
  
  jaccard <- if (union_len > 0) intersection_len / union_len else NA_real_
  
  # Store result
  jaccard_results[[i]] <- data.frame(
    run = run_i,
    community = community_i,
    jaccard_index = jaccard,
    intersection = intersection_len,
    sign_length = length(signature_set)
  )
}

# Combine all results into a single dataframe
jaccard_df <- do.call(rbind, jaccard_results) %>%
   mutate(perc_sign = intersection/sign_length * 100)

write_tsv(jaccard_df, './results/shuffling_jaccard_df.tsv')
```

### Boxplot similarity
```{r}
jaccard_df <- read_tsv('./results/shuffling_jaccard_df.tsv')

jaccard_df %>%
  separate(run, sep = '_', into = c('perc', 'iter')) -> jaccard_df

jaccard_df$perc <- gsub('perc', '%', jaccard_df$perc)

jaccard_df$community <- gsub('community_', 'CL', jaccard_df$community)

group_colors <- c('CL1' = "grey",
                  'CL2' = "#59BDB2", 
                  'CL3' = "#E7DBAE", 
                  'CL4' = "#F4A2A4", 
                  'CL5' = "#E94587", 
                  'CL6' = "#9CCEE3",
                  'CL7' = "#6B2B68")


jaccard_df$perc <- factor(jaccard_df$perc, levels = c('0%', '5%', '10%', '15%', '25%', '50%', '75%', '100%'))

jaccard_df %>%
  ggplot(aes(x = perc, y = intersection/sign_length * 100, fill = community))+
  geom_boxplot() +
  scale_fill_manual(values = group_colors) +
  xlab('') +
  ylab('% of signatures genes found') +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 0, vjust = .5)) -> boxplot_intesection
```

## Supplementary Figure
```{r}
plot_list <- list(boxplot_intesection, 
                  signatures_num,
                  boxplot_pvalues_bins, 
                  plot_classification)

cowplot::plot_grid(plotlist = plot_list, ncol = 2, 
                   rel_widths = c(c(2,1.5), c(1,1)), align = 'h',
                   labels = LETTERS[1:4]) 

 # ggsave('../Figures/shuffling_validation.pdf', dpi = 300,
 #         width = 8, height = 6, device = 'pdf')
```

**Supplementary Figure. Randomization analysis result. A)** Percentage of genes from the original PatientProfiler run recovered in control signatures across across 50 randomization runs at increasing proportion of shuffled genes (5, 10, 15, 25, 50, 75, 100%) in CPTAC transcriptomics data. **B)** Number of control signatures enriching TCGA patients across proportions of randomized genes. **C)** Prognostic value of control signatures evaluated via Kaplan-Meier analysis. Each boxplot summarizes the adjusted p-values across runs for different shuffling percentages. The horizontal red dashed line marks the 0.05 significance threshold. Background shading indicates p-value ranges. **D)** Similarity between CPTAC and TCGA patient subtype distributions, defined by average cosine similarity across runs. The horizontal red dashed line marks the average cosine similarity of original PatientProfiler run.  
